---
title: "Figure 2B. Change Maps Comparing Differences between Historic & Future SDM"
author: "Madi Calbert"
date: last-updated
output: html_document
---

# Comprehensive Model (Nick's Model)

This model is a SDM for the 7 species across all 12 months. 

We have historic and future binary .tif files where 0 = absent and 1 = presence for each species for all 12 months. 

So when looking at change across historic to future scenarios. We will consider: 
- Loss: 1 → 0
- Gain: 0 → 1
- Stable presence: 1 → 1
- No detection: 0 → 0

```{r}
#Load packages
library(terra)       # raster handling
library(tidyterra)   # geom_spatraster for ggplot
library(ggplot2)     # mapping
library(here)        # paths, since you're already using it
library(tigris) 
library(sf)
library(dplyr)
```

Let's try for one species first, a_menziesii

```{r}
spp <- "a_menziesii"
months <- tolower(month.abb)   # "jan","feb",...,"dec"
```

Load in the historic .tif files
```{r}
# base folders for this species
base_dir <- here("data/maxent_outputs", spp, "lowFilter", "monthly_dist_p10")
hist_dir <- file.path(base_dir, "historic")

# 12 historic files in calendar order
hist_files <- file.path(hist_dir, paste0(months, "_", spp, "_2000_2023.tif"))
stopifnot(all(file.exists(hist_files)))  # safety check

r_hist <- rast(hist_files)
names(r_hist) <- months
r_hist
```

Read in future scenario to compare grids: 
```{r}
fut_dir_45  <- file.path(base_dir, "MIROC45")
fut_files_45 <- file.path(fut_dir_45, paste0(months, "_", spp, "_MIROC45_2070_2099.tif"))
stopifnot(all(file.exists(fut_files_45)))

r_fut_45 <- rast(fut_files_45)
names(r_fut_45) <- months

# Ensure same grid; resample future to historic if needed
if (!compareGeom(r_hist, r_fut_45, stopOnError = FALSE)) {
  r_fut_45 <- resample(r_fut_45, r_hist, method = "near")
}

```

We’ll convert each month’s historic/future pair into a 0–3 code:
- 0 = No detection (both),
- 1 = Loss, 
- 2 = Gain, 
- 3 = Stable presence 


And remember that:
- Loss: 1 → 0
- Gain: 0 → 1
- Stable presence: 1 → 1
- No detection: 0 → 0

```{r}
# Codes: 1=Loss (1→0), 2=Gain (0→1), 3=Stable presence (1→1), 0=No detection (0→0)
calc_change_pair <- function(h, f) {
  ifel(is.na(h) | is.na(f), NA,
       ifel(h == 1 & f == 0, 1,
       ifel(h == 0 & f == 1, 2,
       ifel(h == 1 & f == 1, 3, 0))))
}
```

```{r}
# Make an empty stack with same grid & 12 layers
r_change_45 <- rast(r_hist)
r_change_45[] <- NA
```

```{r}
months <- tolower(month.abb)   # ensure we keep names aligned

for (i in seq_along(months)) {
  r_change_45[[i]] <- calc_change_pair(r_hist[[i]], r_fut_45[[i]])
}
names(r_change_45) <- months
```

```{r}
levels(r_change_45) <- list(data.frame(
  ID    = c(0, 1, 2, 3),
  class = c("No detection (both)",
            "Loss (hist→MIROC45)",
            "Gain (hist→MIROC45)",
            "Stable presence (both)")
))
```

```{r}
# change for one pair (returns a SpatRaster)
calc_change_pair <- function(h, f) {
  terra::ifel(is.na(h) | is.na(f), NA,
       terra::ifel(h == 1 & f == 0, 1,
       terra::ifel(h == 0 & f == 1, 2,
       terra::ifel(h == 1 & f == 1, 3, 0))))
}

# build list of monthly change rasters
chg_list <- lapply(seq_along(tolower(month.abb)), function(i) {
  calc_change_pair(r_hist[[i]], r_fut_45[[i]])
})

# stack them into a single 12-layer SpatRaster
r_change_45 <- terra::rast(chg_list)
names(r_change_45) <- tolower(month.abb)

# quick check: do we have 0/1/2/3 values now?
terra::freq(r_change_45[["mar"]])

```


```{r}
# treat classes as discrete categories for plotting
r_change_45_fac <- as.factor(r_change_45)

# colors + labels (Loss=red, Gain=green, Stable=yellow, NoDetection=grey)
pal <- c("0" = "grey80",   # No Detection 
         "1" = "#B22E2E",  # Loss
         "2" = "#2E8B57",  # Gain
         "3" = "#F6D26D")  # Stable
lab <- c("0" = "No Detection (both)",
         "1" = "Loss (hist→MIROC45)",
         "2" = "Gain (hist→MIROC45)",
         "3" = "Stable Presence (both)")

# plot
months <- tolower(month.abb)

p45 <- ggplot() +
  geom_spatraster(data = r_change_45_fac, na.rm = TRUE) +   # NA=transparent over CA polygon
  facet_wrap(~lyr, ncol = 4,
             labeller = as_labeller(setNames(month.abb, months))) +
  scale_fill_manual(name = "Distribution change",
                    values = pal, breaks = names(pal), labels = lab,
                    na.translate = FALSE) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 11) +                               # clean background
  theme(strip.text = element_text(face = "bold", hjust = 0),
        legend.position = "right") +
  labs(
    title = "Distribution Change for A. menziesii"
  )

p45
```

Get the great valley ecoregion shapefile
```{r}
# --- paths / inputs ---
gv_path  <- here::here("data/Vegetation_-_Great_Valley_Ecoregion_[ds2632].geojson")
scenario <- "MIROC45"  # or "MIROC85"

# --- helper: get target CRS from raster, fallback to EPSG:3310 ---
crs_hist_wkt <- terra::crs(r_hist)
target_crs <- tryCatch(sf::st_crs(crs_hist_wkt),
                       error = function(e) sf::st_crs(3310))
if (is.na(target_crs)) target_crs <- sf::st_crs(3310)

# --- read geojson; set source CRS if missing; transform BEFORE union ---
sf_use_s2(TRUE)  # safe while still in lon/lat
gv_sf <- st_read(gv_path, quiet = TRUE)

if (is.na(st_crs(gv_sf))) {
  message("GeoJSON has no CRS; assuming EPSG:4326 (lon/lat).")
  st_crs(gv_sf) <- 4326
}

gv_sf <- gv_sf |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_cast("MULTIPOLYGON", warn = FALSE) |>
  st_make_valid() |>
  st_transform(target_crs)

# --- dissolve to one polygon (use planar ops) ---
sf_use_s2(FALSE)
gv_union_sf <- gv_sf |>
  mutate(one = 1L) |>
  group_by(one) |>
  summarise(.groups = "drop") |>
  st_make_valid()

gv_poly <- terra::vect(gv_union_sf)

# --- build the scenario stack with 'no detection' removed ---
change_stack <- if (scenario == "MIROC45") r_change_45 else r_change_85
fut_stack    <- if (scenario == "MIROC45") r_fut_45    else r_fut_85

mask_any  <- (r_hist + fut_stack) > 0
change_fac <- as.factor( terra::mask(change_stack, mask_any, maskvalues = 0) )

# --- crop + mask to Great Valley ---
change_cv <- terra::mask( terra::crop(change_fac, gv_poly), gv_poly)

# --- plot (loss=red, gain=green, stable=yellow; grey valley background) ---
pal <- c("1"="#B22E2E","2"="#2E8B57","3"="#F6D26D")
lab <- c("1"=paste0("Loss (hist→", scenario, ")"),
         "2"=paste0("Gain (hist→", scenario, ")"),
         "3"="No change")

months <- tolower(month.abb)
p <- ggplot() +
  geom_sf(data = gv_union_sf, fill = "grey90", color = "grey80", linewidth = 0.2) +
  geom_spatraster(data = change_cv, na.rm = TRUE) +
  facet_wrap(~lyr, ncol = 4,
             labeller = as_labeller(setNames(month.abb, months))) +
  scale_fill_manual("Dist change", values = pal,
                    breaks = names(pal), labels = lab, na.translate = FALSE) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 11) +
  theme(strip.text = element_text(face = "bold", hjust = 0),
        legend.position = "right")

p

```



# Aggregate Model (Eco-relevant Model)

This model is an SDM for each species aggregated across eco-relevant months or months where the species is known to germinate or bloom based on lit review. 

```{r}
# --- Aggregate (eco-relevant) change maps for all species --------------------
# Keeps 0 = "No detection (both)"; facets by species; clipped to Great Valley.

library(terra)
library(tidyverse)
library(dismo)
library(tidyterra)
library(ggplot2)
library(sf)
library(here)

# Species and file root (matches your screenshot/naming)
species <- c("a_menziesii","c_pungens","l_pentachaeta",
             "p_arborea","p_ciliata","a_polycarpa","c_lasiophyllus")
agg_root <- here("data/agg_models_072925")

# Read & dissolve the Great Valley ecoregion (once)
gv_path <- here("data/Vegetation_-_Great_Valley_Ecoregion_[ds2632].geojson")
sf_use_s2(TRUE)
gv_raw <- st_read(gv_path, quiet = TRUE)
if (is.na(st_crs(gv_raw))) st_crs(gv_raw) <- 4326
gv_raw <- gv_raw |>
  st_zm(drop = TRUE, what = "ZM") |>
  st_cast("MULTIPOLYGON", warn = FALSE) |>
  st_make_valid()
sf_use_s2(FALSE)
gv_union <- gv_raw |>
  mutate(one = 1L) |>
  group_by(one) |>
  summarise(.groups = "drop") |>
  st_make_valid()

# Change coder: 0=no detect, 1=Loss, 2=Gain, 3=Stable
calc_change_pair <- function(h, f) {
  terra::ifel(is.na(h) | is.na(f), NA,
       terra::ifel(h == 1 & f == 0, 1,
       terra::ifel(h == 0 & f == 1, 2,
       terra::ifel(h == 1 & f == 1, 3, 0))))
}

# Build one eco-relevant change layer for a species+scenario (keeps zeros)
change_layer_for <- function(spp, scenario = c("MIROC45","MIROC85")) {
  scenario <- match.arg(scenario)
  p10_dir  <- file.path(agg_root, spp, "lowFilter", "p10")
  f_hist   <- file.path(p10_dir, sprintf("p10_%s_2000_2023_ecorelevant.tif", spp))
  f_fut    <- file.path(p10_dir, sprintf("p10_%s_%s_2070_2099_ecorelevant.tif", spp, scenario))
  stopifnot(file.exists(f_hist), file.exists(f_fut))

  r_hist <- terra::rast(f_hist)
  r_fut  <- terra::rast(f_fut)
  if (!terra::compareGeom(r_hist, r_fut, stopOnError = FALSE)) {
    r_fut <- terra::resample(r_fut, r_hist, method = "near")
  }

  # Transform GV to the raster CRS
  crs_wkt   <- terra::crs(r_hist)
  target_crs <- tryCatch(sf::st_crs(crs_wkt), error = function(e) sf::st_crs(3310))
  if (is.na(target_crs)) target_crs <- sf::st_crs(3310)
  gv_t  <- sf::st_transform(gv_union, target_crs)
  gv_tv <- terra::vect(gv_t)

  chg <- calc_change_pair(r_hist, r_fut)                   # 0..3 (keeps 0)
  chg <- terra::mask(terra::crop(chg, gv_tv), gv_tv)      # clip to GV
  chg <- as.factor(chg)                                    # categorical
  names(chg) <- spp
  chg
}

# Build stacks across species for each scenario
chg_45 <- terra::rast(lapply(species, change_layer_for, scenario = "MIROC45"))
chg_85 <- terra::rast(lapply(species, change_layer_for, scenario = "MIROC85"))

# Transform GV for plotting in each CRS
gv45 <- st_transform(gv_union, st_crs(terra::crs(chg_45)))
gv85 <- st_transform(gv_union, st_crs(terra::crs(chg_85)))

# Palette + labels
pal <- c("0"="grey80","1"="#B22E2E","2"="#2E8B57","3"="#F6D26D")
labs <- c("0"="Unsuitable Habitat",
           "1"="Habitat Loss",
           "2"="Habitat Gain",
           "3"="No Change")
facet_labs <- setNames(gsub("_", " ", species), species)
```


```{r}
facet_labs <- c(
  "a_menziesii"    = "A. menziesii",
  "c_pungens"      = "C. pungens",
  "l_pentachaeta"  = "L. pentachaeta",
  "p_arborea"      = "P. arborea",
  "p_ciliata"      = "P. ciliata",
  "a_polycarpa"    = "A. polycarpa",   
  "c_lasiophyllus" = "C. lasiophyllus"
)

p_45 <- ggplot() +
  geom_sf(data = gv45, fill = "grey90", color = "grey80", linewidth = 0.2) +
  geom_spatraster(data = chg_45, na.rm = TRUE) +
  facet_wrap(~lyr, ncol = 4, labeller = labeller(lyr = facet_labs)) +
  scale_fill_manual(name = NULL, 
                    values = pal, breaks = names(pal),
                    labels = labs, na.translate = FALSE) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 11) +
  theme(strip.text = element_text(face = "italic", 
                                  hjust = 0, vjust = 1,
                                  size = 7),
        legend.position = "bottom",
        # legend sizing
        legend.text  = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height = unit(6, "pt"),
        legend.key.width  = unit(10, "pt"),
        legend.spacing.y  = unit(2, "pt"),
        plot.title = element_text(margin = margin(b = 5))) +
  labs(title = "Current to MIROC45 Distribution")

p_85 <- ggplot() +
  geom_sf(data = gv85, fill = "grey90", color = "grey80", linewidth = 0.2) +
  geom_spatraster(data = chg_85, na.rm = TRUE) +
  facet_wrap(~lyr, ncol = 4, labeller = labeller(lyr = facet_labs)) +
  scale_fill_manual(name = NULL, 
                    values = pal, breaks = names(pal),
                    labels = labs, na.translate = FALSE) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 11) +
  theme(strip.text = element_text(face = "italic", 
                                  hjust = 0, vjust = 1, 
                                  size = 7),
        legend.position = "bottom",
        # legend sizing
        legend.text  = element_text(size = 7),
        legend.title = element_text(size = 8),
        legend.key.height = unit(6, "pt"),
        legend.key.width  = unit(10, "pt"),
        legend.spacing.y  = unit(2, "pt"),
        plot.title = element_text(margin = margin(b = 5))) +
  labs(title = "Current to MIROC85 Distribution")


pad_legend <- theme(
  plot.margin       = margin(t = 5, r = 5, b = 16, l = 5),  # more room below
  legend.box.margin = margin(t = 2, r = 2, b = 8,  l = 2)   # padding around legend
)

p_45 <- p_45 + pad_legend
p_85 <- p_85 + pad_legend


p_45
p_85
```

```{r}
# Choose the font family 
tnr_family <- "Times"  

facet_labs <- c(
  "a_menziesii"    = "A. menziesii",
  "c_pungens"      = "C. pungens",
  "l_pentachaeta"  = "L. pentachaeta",
  "p_arborea"      = "P. arborea",
  "p_ciliata"      = "P. ciliata",
  "a_polycarpa"    = "A. polycarpa",   
  "c_lasiophyllus" = "C. lasiophyllus"
)

p_45 <- ggplot() +
  geom_sf(data = gv45, fill = "grey90", color = "grey80", linewidth = 0.2) +
  geom_spatraster(data = chg_45, na.rm = TRUE) +
  facet_wrap(~lyr, ncol = 4, labeller = labeller(lyr = facet_labs)) +
  scale_fill_manual(name = NULL, values = pal, breaks = names(pal),
                    labels = labs, na.translate = FALSE) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 11, base_family = tnr_family) +
  theme(
    text = element_text(family = tnr_family),
    strip.text = element_text(family = tnr_family, face = "italic",
                              hjust = 0, vjust = 1, size = 8),
    legend.position = "bottom",
    legend.text  = element_text(family = tnr_family, size = 8),
    legend.title = element_text(family = tnr_family, size = 8),
    legend.key.height = unit(6, "pt"),
    legend.key.width  = unit(10, "pt"),
    legend.spacing.y  = unit(2, "pt"),
    plot.title = element_text(family = tnr_family, margin = margin(b = 5))
  ) +
  labs(title = "Current to MIROC RCP 4.5 Distribution", size = 11)

p_85 <- ggplot() +
  geom_sf(data = gv85, fill = "grey90", color = "grey80", linewidth = 0.2) +
  geom_spatraster(data = chg_85, na.rm = TRUE) +
  facet_wrap(~lyr, ncol = 4, labeller = labeller(lyr = facet_labs)) +
  scale_fill_manual(name = NULL, values = pal, breaks = names(pal),
                    labels = labs, na.translate = FALSE) +
  coord_sf(expand = FALSE) +
  theme_void(base_size = 11, base_family = tnr_family) +
  theme(
    text = element_text(family = tnr_family),
    strip.text = element_text(family = tnr_family, face = "italic",
                              hjust = 0, vjust = 1, size = 8),
    legend.position = "bottom",
    legend.text  = element_text(family = tnr_family, size = 8),
    legend.title = element_text(family = tnr_family, size = 8),
    legend.key.height = unit(6, "pt"),
    legend.key.width  = unit(10, "pt"),
    legend.spacing.y  = unit(2, "pt"),
    plot.title = element_text(family = tnr_family, margin = margin(b = 5))
  ) +
  labs(title = "Current to MIROC RCP 8.5 Distribution")

pad_legend <- theme(
  plot.margin       = margin(t = 5, r = 5, b = 16, l = 5),
  legend.box.margin = margin(t = 2, r = 2, b = 8,  l = 2)
)

p_45 <- p_45 + pad_legend
p_85 <- p_85 + pad_legend

p_45
p_85

```





